Challenges and Solutions of Developing and Implementing a Novel Desktop-as-a-Service

### 2024-05-07 paper.tex
More proof read, shortening and references

Several works in literature have analyzed the latency of Cloud gaming services that have, in some aspects, similar requirements to DaaS offerings. According to Lampe et al.~\cite{lampe2014assessing}, the latency of locally executed games is compromised of the following components:

\begin{itemize}
	\item Input lag caused by the controller device (e.g., mouse or keyboard).
	\item CPU time for processing the user input and the program itself.
	\item GPU time of the graphics processor for rendering the next frame.
	\item Time required for delivering the frame into the frame buffer.
	\item LCD response time (time required to display the frame on the screen).
\end{itemize}

In cloud gaming scenarios, the following components increase latency further:

\begin{itemize}
	\item Transfer time (upstream) for sending the user input to the service provider.
	\item Time required for capturing the frames and encoding it as a video stream.
	\item Transfer time (downstream) for sending the video stream to the client.
	\item Time required for decoding the video stream back into a frame.
\end{itemize}
% TODO: Mention hardware-accelerated framebuffer protocols like \cite{hanimplementation} \cite{darcy2022latency}
% TODO: More scientific papers: \cite{di2021network} \cite{carrascosa2022cloud}
For every deployment scenario or cloud service category possible, users expect a latency value that is not annoying when interacting with the service and its applications. Choy et al.~\cite{ChoyWongSimonRosenberg2012}, Claypool et al.~\cite{claypool2010latency}, and Jarschel et al.~\cite{jarschel2011evaluation} consider a maximum tolerable latency, based on subjective tests, being around 100\,ms. This relatively low value sets the demand for a fast network interaction between client and service and limits their maximum distance. As mentioned in Clincy et al.~\cite{clincy2013subjective}, The former cloud gaming service provider OnLive specified in the year 2010 that the distance between the service provider and the consumer should not be greater than 1000 miles (approx. 1600\,km) because otherwise, the round-trip communications delay time between will be too long for video games.

@@

\subsection{Performance}
To describe the mission-critical performance aspects of the discussed system,
we initially provide a brief overview of potential challenges
by outlining bottlenecks and other relevant problem areas.
This is done in the context of the underlying use case model of DaaS systems.
Subsequently, we give a more comprehensive description
of our specific design decisions and tailored solutions.

Performance, from our viewpoint, stands out as the most crucial system property,
implying the influence of other  relevant properties,
such as user experience, system responsiveness, and overall efficiency.
Consequently, we encountered multifaceted challenges
that potentially intersected across several layers within our architecture.

Hence, every component from the core infrastructure
to all backend services and user-facing elements,
must meet individual and domain-specific requirements
to contribute to the overall performance goals of the entire system.
Thus, we outline key limiting factors that hinder performance,
along with our experiences and the chosen mitigation strategies.

Given the absence of a generalized measurement system,
we ranked these factors based on their magnitude of impact,
drawing from our experiences obtained during the project.

@@

Regarding Infrastructure-as-a-Service (IaaS),
our solution generally offers the flexibility
to use either virtual machines or containers.
Scientific publications discussing relevant performance aspects of such systems
highlight the apparent advantage of a lightweight container solution
over virtual machines when identical programs are tested in both environments
~\cite{felter2015updated,potdar2020performance,seo2014performance}.

@@

Regarding Platform-as-a-Service (PaaS),
studies show that there is no dramatic difference in performance
when derivatives of operating systems are compared to each other
~\cite{balen2020performance,boras2020performance}.
Nevertheless, some derivatives are more lightweight and performant.
In contrast, comparing operating systems of different types
reveals more obvious results, particularly when considering
low-level operating system functionality implementations
such as interrupt handling, memory allocation, and boot time
~\cite{sergeev2022docker,sulaiman2021comparison}.
Additional performance potentials can be leveraged
in cases where a Windows virtual machine can be replaced
with a container-based solution integrating wine
~\cite{huang2012performance}.  % win inside vm

@@

Application-specific requirements can be managed to some extent
through protocol-inherent tuning parameters and compression algorithms.
Our solution, therefore, offers a range of tuning options
to adapt to different use cases
and empowering users to choose their own tailored solutions.

We utilize the Guacamole desktop gateway implementation,
supporting widely used frame-buffer protocols like VNC and RDP,
along with relevant tuning options.
Guacamole also enables audio support and integration with external devices
through a JavaScript-based web client (guacamole.js).
As depicted in figure~\ref{websocket_proxy},
the proxy component receives HTTP requests from the guacamole.js web client
and is forwarded by the backend by using regular WebSocket opcodes (Guacamole protocol).
Such communication is forwarded to the Guacamole daemon (guacd)
which in turn communicates to each instance using the frame-buffer protocol of choice (VNC or RDP).

However, integrating Guacamole into the system
requires consideration of additional factors besides the frame-buffer protocols itself.
Authentication mechanisms, for example, are essential in a DaaS context,
as users expect seamless authentication across the whole system
without multiple logins per session.
Additionally, safeguarding sensitive information like user passwords is crucial
for maintaining system security and user trust.
To mitigate the risk of unintended system manipulation, Guacamole was integrated
using a proxied approach and WebSockets.
This ensures that the credentials of a generated instance remain secret to the host system
while still providing authorized, passwordless access to the hosted application.

@@

\subsubsection{Internal processing and Communication}
Another critical aspect directly related to performance and responsiveness
is how data is internally processed, stored, and communicated
among all involved components and instances.
This includes metadata collected and processed by essential components
and involves the direct exchange of raw data with object instances.

Ensuring a satisfying user experience in a DaaS system
implies specific components in a host system.
This includes access to all backend services
used to create containers and virtual machines,
local or remote databases, authentication services, filesystems, and network devices.
Additionally, a web server is integrated to provide relevant API endpoints
for frontend communication and WebSocket endpoints for viewer communication.

Generally spoken, all of these components can potentially become bottlenecks
when handling high throughputs or large amounts of concurrent applications
requiring low latency access, especially if all necessary backend components
are served from the same hardware.
Architecturally, all relevant components
must be designed for distribution across different hosting systems.
However, retaining specific components directly on the host system
may still be required to ensure low-latency access.

From our observations, components requiring real-time and low-latency performance,
should physically be kept as close to the core hosting system as possible
as shown in figure~\ref{perf_internal_communication}.
This includes all data and communication relevant to
audio, video, mouse, and keyboard transmission,
as well as all backend services for hosting containers and virtual machines.
Ideally, these components should be hosted natively on the host system
for virtual network and hardware access
or at least be connected to the same physical network
to minimize network latencies.

Regarding other components, such as database access,
authentication providers, and filesystems,
maintaining a close distance from the host is advisable,
but this depends mainly on the total number of users
as well as their specific use cases.
On smaller scales, keeping these components
in close physical proximity to the host remains prudent.

@@

\subsection{Stability}
Stability in DaaS applications is especially important
for meeting the demands of continuous and time-critical operations
across diverse user scenarios.
Ensuring the stability of all system components during runtime is paramount
to maintain service continuity, handle growing user loads,
and transparently recover from failures
while providing a seamless end-user experience.
These considerations are integral to our architectural concept,
directly impacting the operational integrity of the system
and its suitability for enterprise-level deployment and trust.

This definition of stability is supported
by various architectural and operational strategies,
including component redundancy, node scalability,
and a robust backend service framework facilitating communication
between components and nodes.
These strategies, along with their drawbacks and benefits,
will be further explored at a component level
and within the following subsections.

\subsubsection{Scalability of Components}

Scalability in DaaS environments
can be approached through two fundamental methods: horizontal and vertical scaling.
Horizontal scaling, or scaling out, involves adding more machines or instances
to manage increased loads.
In contrast, vertical scaling, or scaling up,
adds more power (CPU, RAM) to existing machines
~\cite{vaquero2011dynamically}.
While horizontal scaling offers flexibility
and is often preferred in cloud environments,
vertical scaling remains important for applications
requiring strong single-thread performance or low latencies.

To comply with both approaches,
we designed the system and its backend services
to be split into logical units that can be distributed across dedicated systems.
Each backend service receives a dedicated configuration
and can be accessed via hostname and port tuples.
By default, all backend services reside on the same host,
with tuples derived from the host configuration.
In a distributed context, services can be distributed to dedicated remote systems
to adhere to horizontal scaling principles.
Relevant tuples can then be derived from any node
using a common database containing meta information.

@@

	      and is, therefore, independent of surrounding components and contexts.
	      It can be distributed to a dedicated system if necessary.

@@

	      such as data replication, clustering, failover mechanisms,
	      and RAID control.

@@

	      but can be easily applied by users using such database-specific features

@@

	      allowing for distribution to any dedicated system.
	      It can also be targeted by standardized load-balancing mechanisms
	      and remains entirely agnostic of surrounding components
	      or calling contexts.
	      The authentication system maintains its credential registry
	      in a separate database, along with a customizable set
	      of user and group privileges.
	      Additionally, it implements the well-defined OAuth 2.0 standard
	      to facilitate remote authentication.
	      a customizable number of redundant systems may be configured.

@@

	      Within our architecture, we utilize the distributed filesystem Ceph,
	      which offers object, block, and file storage capabilities.
	      Redundancy in Ceph is implicitly achieved

@@

	      by automatically replicating data across different devices.
	      Our architecture leverages these capabilities
	      to reliably provide file storage to host systems
	      and all hosted DaaS instances,
	      while also offering block storage to the infrastructural backend services.
	      Hosting DaaS instances directly from the distributed block storage
	      enables migration and redundancy of instances or, in other words, horizontal scaling.

@@

			\draw[<->, thick, black] (-11,-7.5) -- (-2.1,-7.5)node[midway, below, font=\footnotesize] {User-Facing Components};
			\draw[<->, thick, black] (-1.9,-7.5) -- (7,-7.5)node[midway, below, font=\footnotesize] {Service Components};

@@

\subsubsection{Internal Communication}
After discussing the principles of scalability and redundancy,
we finally discuss how communication flows are managed in our distributed environment.
Our proposed solution is rooted in a message-oriented architecture
designed to fulfill DaaS properties for scalability and redundancy,
as previously outlined.

This implies that any action triggered, whether in the frontend on the client side,
backend on the host side, or within hosted instances,
must ensure that the necessary information for execution
is readily available at the time of processing.

We achieve this by storing runtime information in distributed block storage,
relevant configuration files in distributed file storage,
object metadata in a common database,
and by utilizing a decentralized OAuth component for authentication.

This ensures that any change triggered by a node within the DaaS domain
is immediately available to any other node relying in some way
on that particular information change.
Additionally, any extra information needed by a component must be fully contained
within the related request or response.
As shown in figure~\ref{message_routing}, this enables arbitrarily distributed configurations as long as request
and response messages reach their desired target eventually
and as long relevant backend services remain available.

If that is the case, each distributed DaaS node might only contain essential and time-critical components,
including the backend service hosting relevant instances as well as the described
WebSocket proxy connected to the guacamole daemon and the user.
Any other component may reside elsewhere within the DaaS network.

@@

For future work, we plan to do a performance analysis that includes an evaluation of different user behaviors for various popular applications and realistic deployment and utilization scenarios.

>>>>

Several works in literature have analyzed
the latency of Cloud gaming services that have, in some aspects,
similar requirements to DaaS offerings.
According to Lampe et al.~\cite{lampe2014assessing},
the latency of locally executed games is compromised of the following components:
\begin{itemize}
	\item Input lag caused by the controller device (e.g., mouse or keyboard).
	\item CPU time for processing the user input and the program itself.
	\item GPU time of the graphics processor for rendering the next frame.
	\item Time required for delivering the frame into the frame buffer.
	\item LCD response time (time required to display the frame on the screen).
\end{itemize}

In cloud gaming scenarios the following components increase latency further:
\begin{itemize}
	\item Transfer time (upstream) for sending the user input to the service provider.
	\item Time required for capturing the frames and encoding it as a video stream.
	\item Transfer time (downstream) for sending the video stream to the client.
	\item Time required for decoding the video stream back into a frame.
\end{itemize}

For every deployment scenario or cloud service category possible,
users expect a latency value that is not annoying
when interacting with the service and its applications.
Choy et al.~\cite{ChoyWongSimonRosenberg2012},
Claypool et al.~\cite{claypool2010latency},
and Jarschel et al.~\cite{jarschel2011evaluation}
consider a maximum tolerable latency, based on subjective tests,
being around 100\,ms. This relatively low value sets the demand for a fast network interaction
between client and service and limits their maximum distance.
As mentioned in Clincy et al.~\cite{clincy2013subjective},
The former cloud gaming service provider OnLive specified in the year 2010
that the distance between the service provider and the consumer
should not be greater than 1000 miles (approx. 1600\,km) because otherwise,
the round-trip communications delay time between will be too long for video games.
A more in-depth comparison of
commercially available cloud gaming services is given
by Di Domenico et al.~\cite{di2021network}
or by Carrascosa et al.~\cite{carrascosa2022cloud}.
Additionally, implied delays and round-trips can be further reduced
by using more sophisticated transmission protocols
directly interacting with GPU drivers,
such as Moonlight~\cite{hanimplementation}~\cite{darcy2022latency}.

@@

\subsection{Performance}
Performance, from our viewpoint, stands out as the most crucial system property,
implying the influence of other  relevant properties,
such as user experience, system responsiveness, and overall efficiency.
Consequently, we encountered multifaceted challenges
that potentially intersected across several layers within our architecture
and which will be described in the following subsections.

% To describe the mission-critical performance aspects of the discussed system,
% we initially provide a brief overview of potential challenges
% by outlining bottlenecks and other relevant problem areas.
% This is done in the context of the underlying use case model of DaaS systems.
% Subsequently, we give a more comprehensive description
% of our specific design decisions and tailored solutions.
% Hence, every component from the core infrastructure
% to all backend services and user-facing elements,
% must meet individual and domain-specific requirements
% to contribute to the overall performance goals of the entire system.
% Thus, we outline key limiting factors that hinder performance,
% along with our experiences and the chosen mitigation strategies.
% Given the absence of a generalized measurement system,
% we ranked these factors based on their magnitude of impact,
% drawing from our experiences obtained during the project.

@@

Regarding Infrastructure-as-a-Service (IaaS),
our solution generally offers the flexibility
to use either virtual machines or containers.
Scientific publications discussing relevant performance aspects of such systems
highlight the apparent advantage of a lightweight container solution
over virtual machines when identical programs are tested in both environments
~\cite{felter2015updated,potdar2020performance}.

@@

Regarding Platform-as-a-Service (PaaS),
studies show that there is no dramatic difference in performance
when derivatives of operating systems are compared to each other
~\cite{balen2020performance,boras2020performance}.
Nevertheless, some derivatives are more lightweight and performant.
In contrast, comparing operating systems of different types
reveals more obvious results, particularly when considering
low-level operating system functionality implementations
such as interrupt handling, memory allocation, and boot time
~\cite{sulaiman2021comparison}.
Additional performance potentials can be leveraged
in cases where a Windows virtual machine can be replaced
with a container-based solution integrating wine
~\cite{huang2012performance}.

@@

Application-specific requirements can be managed to some extent
through protocol-inherent tuning parameters and compression algorithms.
Our solution, therefore, offers a range of tuning options
to adapt to different use cases on a per-user basis.

We utilize the Guacamole desktop gateway implementation,
supporting widely used frame-buffer protocols like VNC and RDP,
along with relevant tuning options.
Guacamole also enables audio support and integration with external devices
through a JavaScript-based web client (guacamole.js).
As depicted in figure~\ref{websocket_proxy},
the proxy component receives HTTP requests from guacamole.js
and is forwarded by the backend through WebSocket opcodes (Guacamole protocol).
These opcodes are then transmitted to the Guacamole daemon (guacd)
which in turn communicates to each instance
by using regular frame-buffer protocols (VNC or RDP).

However, integrating Guacamole into the system
requires consideration of additional factors besides the frame-buffer protocols itself.
Authentication mechanisms, for example, are essential in a DaaS context,
as users expect seamless authentication across the whole system
without multiple logins.
Additionally, safeguarding sensitive information like user passwords is crucial
for maintaining system security and user trust.
To mitigate the risk of unintended system manipulation, Guacamole was integrated
using a proxied approach and WebSockets.
This ensures that the credentials of a generated instance remain secret to the host system
while still providing authorized, passwordless access to the hosted application.

@@

\subsubsection{Internal processing and Communication}
Another critical aspect directly related to performance and responsiveness
is how data is internally processed, stored, and communicated.
This includes metadata collected and processed
and involves data exchange with all relevant backend services or instances.

Ensuring a satisfying user experience in a DaaS system
implies specific components to be available.
This includes access to all infrastructural backend services,
a databases, authentication services, filesystems, and network devices.
Additionally, a web server is integrated to provide relevant API endpoints
for frontend communication and WebSocket endpoints for the viewer.

Generally spoken, all of these components can potentially become bottlenecks
when handling high throughputs or large amounts of concurrent applications
and is especially true if all components are served from the same hardware.
Architecturally, all relevant components
must therefore be designed for distribution across different hosts.
However, keeping specific components directly on the host system
may still be required to ensure low-latency access.

From our observations, components requiring real-time and low-latency performance,
should physically be kept as close to the core hosting system as possible
as shown in figure~\ref{perf_internal_communication}.
This includes all data and communication relevant to
audio, video, mouse, and keyboard transmission,
as well as all backend services for hosting containers and virtual machines.
Ideally, these components should be hosted natively on the host system
for virtual network and hardware access
or at least be connected to the same physical network
to minimize network latencies.
All other components might be distributed to dedicated systems
if large user amounts and specific use cases have to be handled
but keeping them in close physical proximity
to the host remains prudent in smaller scales.

Allthough we can not yet provide a full-fledged performance analysis,
we already measured promising results in our local test environment.
Depending on the exact setup,
the solution provides
reasonable round-trip times and delays for each communication direction
visualized in figure~\ref{perf_internal_communication}.
In a minimal setup, we measured delays of around 10ms for client communication,
below 1ms for internal backend communication
but up to 1000ms for instance communication.
But these numbers are only rough estimates
as certain optimization concepts remain to be implemented.

@@

\subsection{Stability}
Stability in DaaS applications is especially important
for meeting the demands of continuous and time-critical operations
across diverse user scenarios.
Ensuring the stability of all system components during runtime is paramount
to maintain service continuity, handle growing user loads,
and transparent failure recovery while providing a seamless end-user experience.
These considerations are integral to our architectural concept,
directly impacting the operational integrity of the system
and its suitability for enterprise-level deployment and trust.

Therefore, key strategies, such as component redundancy, node scalability
as well as the internal communication concept
will be described in the following subsections.

\subsubsection{Scalability of Components}
Scalability in DaaS environments
can be approached through two fundamental methods: horizontal and vertical scaling.
Horizontal scaling (scale out), involves adding more machines or instances
to manage increased loads.
In contrast, vertical scaling (scale up),
adds more power (CPU, RAM) to existing machines
~\cite{vaquero2011dynamically}.
While horizontal scaling offers flexibility
and is often preferred in cloud environments,
vertical scaling remains important for applications
requiring strong single-thread performance or low latencies.

To comply with both approaches,
we designed the system and its backend services
to be split into logical units that can be distributed across dedicated systems.
Each backend service receives a dedicated configuration
and can be accessed via hostname and port tuples.
By default, all backend services reside on the same host.
In a distributed context, services can be distributed to dedicated remote systems
to adhere to horizontal scaling principles.

@@

	      and is, therefore, independent of surrounding components and contexts
	      and can be externalized to a dedicated systems.

@@

	      such as data replication, clustering, failover mechanisms or RAID control.

@@

	      but can be easily applied by using such database-specific features

@@

	      also allowing for distribution to dedicated systems.
	      It can be targeted by standardized load-balancing mechanisms
	      and remains entirely agnostic of calling contexts.
	      The system maintains its credential registry in a separate database,
	      along with customizable user and group privileges and implements
	      the well-defined OAuth 2.0 standard to facilitate remote authentication.

@@

	      In our architecture, we utilize the distributed filesystem Ceph,
	      which offers means of object, block, and file storage.
	      Redundancy is implicitly achieved

@@

	      by automatically replicating data across multiple systems.
	      Such benefits are leveraged by providing reliable file storage to all instances,
	      and by providing block storage to the infrastructural backend services.
	      By doing that, migration and redundancy of instances is achieved
	      or, in other words, horizontal scaling.

@@

			\draw[<->, thick, black] (-11,-7.5) -- (-2.1,-7.5) node[midway, above, yshift=-0.5cm, font=\footnotesize] {User-Facing Components};
			\draw[<->, thick, black] (-1.9,-7.5) -- (7,-7.5) node[midway, above, yshift=-0.5cm, font=\footnotesize] {Service Components};

@@

\subsubsection{Internal Communication}
After discussing the principles of scalability and redundancy,
we finally describe how communication flows are managed within our system.
Our solution is rooted in a message-oriented architecture
designed to fulfill DaaS properties for scalability and redundancy,
as previously outlined.

This implies that any action triggered, whether in the frontend on the client side,
in the backend on the host side, or within hosted instances,
must ensure that all necessary information is readily available at the time of processing.

We achieve this by storing runtime information in distributed block storage,
relevant configuration files in distributed file storage,
object metadata in a common database,
and by utilizing a decentralized authentication component.

This ensures that any change triggered by a node within the DaaS domain
is immediately available to any other node relying in some way
on that particular information change.
Additionally, any extra information needed by a component must be fully contained
within the related request.
As shown in figure~\ref{message_routing}, this enables arbitrarily distributed configurations as long as request
and response messages reach their desired target eventually.

If that is the case, each distributed host system might only contain it's essential components,
including the backend service with all hosted instances
as well as the described proxy component for direct user interaction.
Any other component may reside somewhere in the DaaS network.

@@

For future work,
we plan to replace our temporary solution for internal communication
with a more efficient approach as our preliminary performance tests revealed
a potential bottlenack in that particular area.
We also plan to apply certain optimization strategies
with respect to backend and instance communication.
We further plan to enhance our yet limited performance testing
and will provide a full-fledged performance analysis
of realsitic deployment scnarios and utilization patterns
which include an evaluation of different user behaviors
within various popular applications.
